Notes for Week 8/1/19 - 8/8/19

### 2019-08-02 notes/weekly_notes/8.1.19.NotesForWeek.tex
updated weekly notes; began drafting master document

\begin{document}
\title{Graph Testing}
\author{Alden Green}
\date{\today}
\maketitle

\section{Fixed graph testing.}

Suppose we observe $G = (V,E)$, an undirected graph over $V = [n]$. Let $B$ be the $m \times n$ incidence matrix of $G$, with singular value decomposition $B = U\Lambda^{1/2}V^T$, so that the Laplacian matrix $L = V \Lambda V^T$. For $i \in [n]$, we observe
\begin{equation*}
z_i = \beta_i + \varepsilon_i, \quad \varepsilon_i \overset{\textrm{i.i.d}}{\sim} \mathcal{N}(0,1)
\end{equation*}
Our statistical goal is hypothesis testing. We wish to distinguish
\begin{equation*}
\mathbf{H}_0: \norm{\beta}_2 = 0 \quad \textrm{vs.} \quad \mathbf{H}_a: \norm{\beta}_2 > 0.
\end{equation*}
We will evaluate our performance using the notion of \emph{worst-case error}. For a given ``function'' class $\mathcal{H}$, test function $\phi: \Reals^n \to \set{0,1}$, and $\epsilon > 0$, let
\begin{equation*}
\mathcal{R}_{\epsilon}(\phi;\mathcal{H})= \Ebb_{0}(\phi) + \sup_{\beta \in \mathcal{H}:\norm{\beta}_2 > \epsilon} \Ebb_{\beta}(1 - \phi)
\end{equation*}

\subsection{Discrete Sobolev Classes.}

The first ``function'' class we will consider will be balls in discrete Sobolev norms, with known smoothness. For $s,d$ positive integers, and radius $C_n > 0$, let
\begin{equation*}
\mathcal{S}^{s}_{d}(C_n) = \set{\beta: \norm{D^{(s)}\beta}_2 \leq C_n}
\end{equation*}
where
\begin{equation*}
D^{(s)} = 
\begin{cases}
L^{s/2}, & \text{$s$ even} \\
DL^{(s-1)/2}, & \text{$s$ odd}.
\end{cases}
\end{equation*}
We note that the constraint is equivalent to $\beta^T V \Lambda^s V \beta \leq C_n^2$.
\subsection{Test statistic}
Let $0 = \lambda_0 \leq \lambda_1 \leq \lambda_2 \leq \ldots \leq \lambda_{n - 1}$ denote the ordered eigenvalues of $L$, and let $v_k$ denote the eigenvector corresponding to $\lambda_k$. For a constant $C > 0$ to be specified later, let 
\begin{equation*}
T_C = \sum_{k: \lambda_k^{s} \leq C^2} y_k^2
\end{equation*}
where
\begin{equation*}
y_k = \frac{1}{\sqrt{n}} \dotp{z}{v_k}.
\end{equation*}

We first compute the expectation $\Ebb(T_C)$. Let $\theta \in \Reals^n$ represent the expectation of $(y_k)$, meaning
\begin{equation*}
\theta_k = \frac{1}{\sqrt{n}}\dotp{\beta}{v_k}
\end{equation*}
and let $\Pi_C\theta$ have entries $(\Pi_C \theta)_k = \1(\lambda_k^{s} \leq C^2) \theta_k$. Finally, let $N(C) = \sharp\set{k: \lambda_k^{s} \leq C^2}$.

\begin{lemma}
	\label{lem:expectation}
	For any $\beta \in \Reals^n$,
	\begin{equation}
	\label{eqn:expectation_1}
	\Ebb(T_C) = \frac{N(C)}{n} + \norm{\Pi_C \theta}_2^2 
	\end{equation}
	If additionally $\beta \in \mathcal{H}$, the following lower bound holds:
	\begin{equation}
	\label{eqn:expectation_2}
	\Ebb(T_C) \geq \frac{N(C)}{n} + \frac{\norm{\beta}^2}{n}  - \frac{C_n^2}{nC^{2}}
	\end{equation}
\end{lemma}
\begin{proof}
	We can write
	\begin{align*}
	\Ebb(T_C) & = \sum_{k: \lambda_k^s \leq C} \Ebb(y_k^2) \\
	& = \frac{1}{n}\sum_{k: \lambda_k^s \leq C} \Ebb\bigl(\dotp{\beta}{v_k}^2 + \dotp{\varepsilon}{v_k}^2 + 2 \dotp{\varepsilon}{v_k} \dotp{\beta}{v_k}\bigr) \\
	& = \sum_{k: \lambda_k^s \leq C}\theta_k^2 + \frac{1}{n} \\
	& = \norm{\Pi_C \theta}^2 + \frac{N(C)}{n},
	\end{align*}
	showing \eqref{eqn:expectation_1}. Now, assuming, $\norm{D^{(s)}\beta}_2 \leq C_n$, we can further obtain
	\begin{align*}
	\norm{\Pi_C \theta}^2 & = \norm{\theta}^2 - \sum_{k:\lambda_k^s > C^2} \theta_k^2 \\
	& \geq \norm{\theta}^2 - \frac{1}{C^{2}}\sum_{k:\lambda_k^s > C^2} \theta_k^2 \lambda_k^s \\
	& \geq \norm{\theta}^2 -\frac{1}{nC^2} \beta^T V \Lambda^s V^T \beta \\
	& \geq \norm{\theta}^2  - \frac{C_n^2}{nC^{2}}
	\end{align*}
	and \eqref{eqn:expectation_2} is shown.
\end{proof}

We now turn to computing the variance $\Var(T_C)$.
\begin{lemma}
	\label{lem:variance}
	\begin{equation*}
	\Var(T_C) = \frac{2N(C)}{n^2} + \frac{4 \norm{\Pi_C \theta}_2^2}{n}
	\end{equation*}
\end{lemma}
\begin{proof}
	To begin, we rewrite
	\begin{align*}
	T_C & = \sum_{k:\lambda_k^s \leq C^2} y_k^2 \\
	& = \frac{1}{n} \sum_{k: \lambda_k^s \leq C^2} \dotp{z}{v_k}^2 \\
	& = \frac{1}{n} \sum_{k: \lambda_k^s \leq C^2} z^T v_k v_k^T z \\
	& =: \frac{1}{n} z^T P_{C} z.
	\end{align*}
	where $P_C := \sum_{k: \lambda_k^s \leq C^2} v_k v_k^T$. Therefore $\Var(T_C) = \Var(z^T P_{C} z)/n^2$. We expand $z = \beta + \varepsilon$ to obtain
	\begin{align}
	\Var(z^T P_{C} z) & = \Var((\beta + \epsilon)^T P_{C} (\beta + \epsilon)) \nonumber \\
	& = \Var(\beta^T P_{C} \beta + 2 \epsilon P_{C} \beta + \epsilon^T P_{C} \epsilon) \nonumber \\
	& = 4 \beta^T P_{C} I P_{C} \beta + \Var(\epsilon^T P_{C} \epsilon) + 4 \Cov(\epsilon P_{C} \beta, \epsilon^T P_{C} \epsilon) \nonumber \\
	& = 4 n \norm{\Pi_C \theta}_2^2 + \Var(\epsilon^T P_{C} \epsilon) \label{eqn:variance_1}
	\end{align}
	where the last equality follows from the Gaussianity of $\epsilon$, as
	\begin{equation*}
	\Ebb\bigl( (\epsilon P_{C} \beta) (\epsilon^T P_{C} \epsilon) \bigr) = \sum_{i = 1}^{n} \sum_{j = 1}^{n} \sum_{k = 1}^{n} (P_C\beta)_k (P_C)_{ij} \Ebb(\epsilon_i \epsilon_j \epsilon_k) = 0.
	\end{equation*}
	Also by the Gaussianity of $\epsilon$, $\epsilon^T P_C \epsilon \sim \chi_{N(C)}^2$, and therefore $\Var(\epsilon^T P_{C} \epsilon) = 2N(C)$. Plugging back into \eqref{eqn:variance_1}, we obtain
	\begin{equation*}
	\Var(z^T P_{C} z) = 4 n \norm{\Pi_C \theta}_2^2 + 2N(C) 
	\end{equation*}
	and therefore the desired result is proved.
\end{proof}

We will consider now the test $\phi_C = \mathbf{1}\{T(C) \geq N(C)/n + \tau(b)\}$, where for $b \geq 1$, $\tau(b) = b \sqrt{2N(C)/n^2}$. We first upper bound the type I error.

\begin{lemma}
	\label{lem:type_I_error}
	Under the null hypothesis $\beta = 0$, and for any $C > 0$,
	\begin{equation*}
	\Ebb_{\beta = 0}(\phi_C) \leq \frac{1}{b^2}. 
	\end{equation*}
\end{lemma}
\begin{proof}
	The desired result follows from Chebyshev's inequality,
	\begin{align*}
	\Ebb_{\beta = 0}(\phi) & = \Pbb_{\beta = 0}\bigl(T(C) \geq N(C)/n + \tau(b)\bigr) \\ 
	& = \Pbb_{\beta = 0}\bigl(T(C) - \frac{N(C)}{n} \geq \tau(b)\bigr) \\
	& \leq \Pbb_{\beta = 0}\bigl(\abs{T(C) - \frac{N(C)}{n}} \geq \tau(b)\bigr) \\
	& \leq \frac{\Var_{\beta = 0}(T_C)}{\tau(b)^2} = \frac{1}{b^2}.
	\end{align*}
\end{proof}
The calculation for the type II error will be slightly more involved.
\begin{lemma}
	\label{lem:type_II}
	Let $b \geq 1$ be fixed. For every $\beta \in \mathcal{S}_d^s(C_n)$ such that
	\begin{equation}
	\label{eqn:type_II}
	\frac{\norm{\beta}^2}{n} \geq 2b\sqrt{2\frac{N(C)}{n^2}} + \frac{C_n^2}{nC^2}
	\end{equation}
	we have that
	\begin{equation*}
	\Ebb_{\beta}(1 - \phi) \leq \frac{2}{b^2} + \frac{2}{b\sqrt{2N(C)}}.
	\end{equation*}
\end{lemma}
\begin{proof}
	Let $\Delta = \Ebb_{\beta}(T_C) - N(C)/n = \norm{\Pi_C \theta}^2$, and observe that by Lemma \ref{lem:expectation} and \eqref{eqn:type_II},
	\begin{equation*}
	\Delta \geq \frac{\norm{\beta}_2^2}{n} - \frac{C_n^2}{nC^2} \geq 2 \tau(b).
	\end{equation*}
	An application of Chebyshev's inequality yields
	\begin{align*}
	\Ebb_{\beta}\bigl(1 - \phi\bigr) & = \Pbb_{\beta}\bigl(T_C \leq N(C)/n + \tau(b)\bigr) \\
	& = \Pbb_{\beta}\bigl(T_C - \Ebb_{\beta}(T_C) \leq \tau(b) - \Delta \bigr) \\
	& \leq \Pbb_{\beta}\bigl(\abs{T_C - \Ebb_{\beta}(T_C)} \leq \Delta - \tau(b) \bigr) \tag{since $\Delta \geq \tau(b)$}	\\
	& \leq \frac{\Var_{\beta}(T_C)}{(\Delta - \tau(b))^2} \\
	& \leq 4\frac{\Var_{\beta}(T_C)}{\Delta^2} \tag{since $\Delta \geq 2\tau(b)$} \\
	& \leq 4\frac{2N(C)/n^2 + \norm{\Pi_C\theta}_2^2/n}{\Delta^2}.
	\end{align*}

	We now handle each summand separately. For the first term, since $\Delta \geq 2 \tau(b)$, we have
	\begin{equation*}
	\frac{2N(C)}{n^2\Delta^2} \leq \frac{1}{2b^2}.
	\end{equation*}

	For the second term, since $\Delta = \norm{\Pi_C\theta}^2$, we have
	\begin{align*}
	\frac{\norm{\Pi_C\theta}_2^2/n}{\Delta^2} & \leq \frac{1}{n\Delta^2} \\
	& \leq \frac{1}{2n\tau(b)} \\
	& = \frac{1}{2b\sqrt{2N(C)}}.
	\end{align*}
\end{proof}

To more explicitly specify the critical radius $\epsilon: \norm{\beta}_2 \geq \epsilon$, we will need to make an assumption on the relation between $N(C)$ and $C$. In particular, let $C = C^*$, where
\begin{equation*}
C^* = c\frac{(C_n n^{s/d})^{4s/(4s + d)}}{n^{s/d}}.
\end{equation*}
for some $c$ constant in $n$. We will assume the following bound on $N(C^*)$:
\begin{enumerate}[label=(A\arabic*)]
	\item Tail decay:
	\label{asmp:tail_decay}
	\begin{equation*}
	N(C^*) \leq (C^*)^{d/s}n
	\end{equation*}
\end{enumerate}

\begin{corollary}
	\label{cor:critical_radius}
	Fix $b \geq 1$, assume \ref{asmp:tail_decay}. Then, for any $C_n > 0$
	\begin{equation*}
	\sup_{\substack{\beta \in \mathcal{S}_d^{s}(C_n), \\ \norm{\beta}_2/n \geq \epsilon}} \Ebb_{\beta}(1 - \phi_{C^{\star}}) \leq \frac{2}{b^2} + \frac{2}{b\sqrt{N(C^{\star})}}
	\end{equation*}
	for every $\epsilon > 0$ such that 
	\begin{equation*}
	\epsilon^2 \geq (2b + 1) \left(\frac{(C_n n^{s/d})^{2d/(4s + d)}}{n}\right).
	\end{equation*}
\end{corollary}
\begin{proof}
	First, we verify that $\epsilon^2 \geq 2b\sqrt{2N(C^*)/n^2} + C_n^2/n(C^*)^2$. By Assumption \ref{asmp:tail_decay} and the choice of $C^{\star}$, we have
	\begin{equation*}
	N(C^*) \leq (C^*)^{d/s}n = (C_n n^{s/d})^{4d/(4s + d)}
	\end{equation*}
	and therefore
	\begin{equation*}
	2b\sqrt{\frac{N(C^*)}{n^2}} \leq \frac{2b}{n} (C_n n^{s/d})^{2d/(4s+d)}
	\end{equation*}
	Moving on to the second term, we have
	\begin{align*}
	\frac{C_n^2}{n(C^{\star})^2} & = \frac{C_n^2n^{2s/d}}{n(C_nn^{s/d})^{8s/(4s+d)}} \\
	& = \frac{C_n^{2d/(4s + d)}n^{(s/d)2d/(4s + d)}}{n} \\
	& = \frac{(C_n n^{s/d})^{2d/(4s + d)}}{n}
	\end{align*}
	and therefore $\epsilon \geq 2b\sqrt{2N(C^*)/n^2} + C_n^2/n(C^*)^2$. As a result, by Lemma \ref{lem:type_II} for any $\beta \in \mathcal{S}^s_d(C_n)$ such that $\norm{\beta}/n \geq \epsilon$, 
	\begin{equation*}
	\Ebb_{\beta}(1 - \phi_{C^{\star}}) \geq \frac{2}{b^2} + \frac{2}{b\sqrt{2N(C^{\star})}}.
	\end{equation*}
\end{proof}

\section{Graph-based Tests over Function Spaces}

Let $\mathcal{D} = [0,1]^d$. Suppose we observe the random design $x_1,\ldots,x_n$ independently sampled from probability measure $P$ over $[0,1]^d$. Additionally, for $i \in [n]$, we observe
\begin{equation*}
z_i = f(x_i) + \varepsilon_i, \quad \varepsilon_i \overset{\textrm{i.i.d}}{\sim} \mathcal{N}(0,1), \quad \varepsilon \perp x
\end{equation*}
where $f \in \mathcal{H} \subseteq L^2(\mathcal{D})$. Our statistical goal is hypothesis testing. We wish to distinguish:
\begin{equation*}
\mathbf{H}_0: \norm{f}_2 = 0 \quad \textrm{vs} \quad \mathbf{H}_{\textrm{a}}: \norm{f}_2 > 0. 
\end{equation*}
We will evaluate our performance using worst-case error: for a given function class $\mathcal{H}$, test function $\phi: \Reals^n \to \set{0,1}$ and $\epsilon> 0$, let
\begin{equation*}
\mathcal{R}_{\epsilon}(\phi; \mathcal{H}) = \Ebb_{f = 0}(\phi) + \sup_{f \in \mathcal{H}: \norm{f}_2 > \epsilon} \Ebb_f(1 - \phi)
\end{equation*}

\subsection{Sobolev Spaces.}

The first type of function class $\mathcal{H}$ we consider will be balls in Sobolev spaces. Let $s,d$ be known, fixed positive integers. For $f: \mathcal{D} \to \Reals$ locally summable, we use the multiindex notation $D^{\alpha}f$ to denote the $\alpha$th-weak partial derivative of $f$ (if one exists). Then, the Sobolev norm is 
\begin{equation*}
\norm{f}_{W^{s,2}(\D)}^2 = \sum_{\abs{\alpha} \leq s} \int_{\mathcal{D}} \norm{D^{\alpha}f}_2^2 \,dx
\end{equation*}
and the corresponding unit ball is $W^{s,2}(\D; L) = \set{f: \norm{f}_{W^{s,2}(\mathcal{D})} \leq L}$.

\subsubsection{Projection-based test statistic.}

For $x,y \in \mathcal{D}$ and radius $r > 0$ to be specified later, let $\eta_r(x,y) = \mathbf{1}(\norm{x - y}/ r \leq 1)$. and let $A$ be the $n \times n$ adjacency matrix with entries $A_{ij} = \eta_r(x_i,x_j)$. Let $B$ be the incidence matrix associated with $A$, and $L = B^TB$ be the corresponding Laplacian matrix. Write $B = U \Lambda^{1/2}V^T$ for the singular value decomposition of $B$. Then $L = V \Lambda V^T$ is the eigendecomposition of $L$ , where $\Lambda$ is a diagonal matrix of eigenvalues with diagonal entries $\lambda_1 \leq \lambda_2 \leq \ldots \leq \lambda_n$, and $V$ is an orthonormal matrix of eigenvectors.

Our test statistic will be the norm of a projection of $z$ onto the subspace spanned by the first eigenvectors of $V$. In particular, for $C > 0$ to be specified later, our test statistic will be
\begin{equation*}
T_C := \sum_{k: \lambda_k^s \leq C^2} y_k^2, \quad y_k = \frac{1}{\sqrt{n}} z^T v_k.
\end{equation*}
For $b \geq 1$, let $\tau(b) = b\sqrt{2N(C)/n^2}$. Then our test will be $\phi_{C} = \mathbf{1}\set{T_C \leq N(C)/n + \tau(b)}$.

In the following result, we prove that for an appropriate choice of threshold $C$ and radius $r$, the test $\phi_C$ achieves (nearly) minimax-optimal rates. Let
\begin{equation*}
C^{\star} = \frac{\left(n^{1 + s/d}r^{d/2 + s}\right)^{4s/(4s + d)}}{n^{s/d}}.
\end{equation*}

\begin{theorem}
	\label{thm:sobolev_testing_rate}
	Let $b \geq 1$ be a fixed constant, and let $s = 1$. For any sequence $r_n$ such that $(\log n/n)^{1/d}r \to \infty$, there exists a constant $c$ which depends only on $d$ and $L$ such that
	\begin{equation}
	\label{eqn:sobolev_testing_rate_1}
	\mathcal{R}_{\epsilon}(\phi_{C^{\star}}; \mathcal{W}^{s,2}(\mathcal{D};L)) \leq c_1\left(\frac{1}{b^2} + \frac{1}{b\sqrt{N(C^{\star})}}\right) + o(1)
	\end{equation}
	for every $\epsilon$ such that
	\begin{equation}
	\label{eqn:sobolev_testing_rate}
	\epsilon^2 \geq c_2 b \left(\frac{(b n^{1 + s/d}r^{d/2 + s})^{2d/(4s + d)}}{n}\right).
	\end{equation}
\end{theorem}

\subsection{Proof of Theorem \ref{thm:sobolev_testing_rate}}
Our goal is to upper bound
\begin{equation*}
\mathcal{R}_{\epsilon}(\phi_{C^{\star}}; W^{1,2}(\mathcal{D};L)) = \Ebb_{f = 0}(\phi_{C^{\star}}) + \sup_{f \in W^{1,2}(\mathcal{D};L): \norm{f}_2 > \epsilon} \Ebb_{f}(1 - \phi_{C^{\star}})
\end{equation*}
By Lemma \ref{lem:type_I_error}, we immediately have $\Ebb_{f = 0}(\phi_{C^{\star}}) \leq 1/b^2$, and we turn to the worst-case type II error. Fix any $f \in W^{1,2}(\mathcal{D};L)$ such that $\norm{f}_2 > \epsilon$, and let $\beta$ be the length-$n$ random vector with $j$th entry $\beta_j = f(x_j)$. Consider the following events:
\begin{enumerate}[label=(E\arabic*)]
	\item 
	\label{event:discrete_sobolev_norm}
	Discrete Sobolev norm of $\beta$: There exists a constant $c_1$ which depends only on $p_{\max}$ and $d$ such that
	\begin{equation*}
	\norm{B^{(s)}\beta}_2^2 \leq c_1 b^2 n^2 r^{d + 2s} 
	\end{equation*}
	\item 
	\label{event:eigenvalue_tail_decay}
	Eigenvalue tail decay:
	\begin{equation*}
	N(C^{\star}) \leq n(C^{\star})^{d/s}.
	\end{equation*}
	\item 
	\label{event:l2_norm}
	$L_2$ norm of $\beta$:
	\begin{equation*}
	\frac{\norm{\beta}_2^2}{n} \geq \frac{1}{2}\epsilon^2.
	\end{equation*}
\end{enumerate}

Each of these events happen with high probability as $n \to \infty$.
\begin{lemma}
	\label{lem:high_probability_events}
	With probability at least $1 - \frac{1}{b^2}$ as $n \to \infty$,
	\begin{equation}
	\label{eqn:discrete_sobolev_norm}
	\norm{B\beta}_2^2 \leq c_1 b^2 n^2 r^{d + 2}.
	\end{equation}

	With probability tending to one as $n \to \infty$,
	\begin{equation}
	\label{eqn:eigenvalue_tail_decay}
	N(C^{\star}) \leq n(C^{\star})^{d/s}.
	\end{equation}

	\textcolor{red}{$L_2$ norm}.
\end{lemma}
In the following subsections, we prove Lemma \ref{lem:high_probability_events}. First, we show that Theorem \ref{thm:sobolev_testing_rate} is a simple consequence of Lemma \ref{lem:high_probability_events} along with our previous work. Let $\mathcal{A}_f$ be the set of $X$ such that \ref{event:discrete_sobolev_norm}, \ref{event:eigenvalue_tail_decay}, and \ref{event:l2_norm} all occur, for the given choice of $f$. We have that for $C_n := c_1bnr^{d/2 + s}$, the event \ref{event:discrete_sobolev_norm} implies $\beta \in \mathcal{S}_d^s(C_n)$. Therefore, by \ref{event:l2_norm},
\begin{equation}
\label{eqn:sobolev_testing_rate_2}
\Ebb_f(1 - \phi_{C^*}) \leq \sup_{\substack{\beta \in \mathcal{S}_d^{s}(C_n), \\ \norm{\beta}_2/n \geq \epsilon^2/2}} \Ebb_{\beta}(1 - \phi_{C^{\star}}) + \Pbb(\mathcal{A}_f).
\end{equation}
Observe the following: 
\begin{itemize}
	\item \textcolor{red}{$C^{\star} = (C_n n^{s/d})^{4s/(4s+d)}/n^{s/d}$},
	\item The event \ref{event:eigenvalue_tail_decay} is exactly \ref{asmp:tail_decay}, and 
	\item For an appropriate choice of $c_2$, by hypothesis
	\begin{equation*}
	\epsilon^2/2 \geq (2\sqrt{2}b + 1) \left(\frac{(C_n n^{s/d})^{2d/(4s + d)}}{n}\right)
	\end{equation*}
\end{itemize}
Therefore we may apply Corollary \ref{cor:critical_radius}, and obtain
\begin{equation*}
\sup_{\substack{\beta \in \mathcal{S}_d^{s}(C_n), \\ \norm{\beta}_2/n \geq \epsilon^2/2}} \Ebb_{\beta}(1 - \phi_{C^{\star}}) \leq \frac{2}{b^2} + \frac{2}{b\sqrt{N(C^{\star})}},
\end{equation*}
and along with \eqref{eqn:sobolev_testing_rate_2} and Lemma \ref{lem:high_probability_events}, this shows
\begin{equation*}
\Ebb_f(1 - \phi_{C^*}) \leq \frac{3}{b^2} + \frac{2}{b\sqrt{N(C^{\star})}} + o(1).
\end{equation*}

\subsection{Proof of \eqref{eqn:discrete_sobolev_norm}}

>>>>

\begin{document}
\title{Notes for Week 8/1/19 - 8/8/19}
\author{Alden Green}
\date{\today}
\maketitle

\section{Bound on Eigenvalue Tail Decay of Neighborhood Graph}
Let $\mathcal{D} = [0,1]^d$. We consider two graphs over data on $\mathcal{D}$.

Suppose we observe the random design $X = x_1, \ldots, x_n$ independently sampled from probability measure $P$ supported on $\mathcal{D}$. For any $r \geq 0$, let the kernel function $\eta_r: \mathcal{D} \times \mathcal{D} \to \Reals_{\geq 0}$ be given by $\eta_r(x,y) = \1(\norm{x - y} \leq r)$. Then, let the \emph{r-neighborhood graph} over $x_1,\ldots,x_n$ be the undirected, unweighted graph $G = (V,E)$, where $V = [n]$ and for $i,j \in [n]$, $(i,j) \in E$ if $\eta_r(x_i, x_j) = 1$.

Let $B$ be the incidence matrix associated with $G$, and let $L = B^TB$ be the corresponding Laplacian. Write $L = V \Lambda V^T$ for the eigen decomposition of $L$, where $V = (v_1 \ldots v_n)$ is an orthonormal matrix with the eigenvectors of $L$ as its columns, and $\Lambda$ is a diagonal matrix with entries $\lambda_1 \leq \lambda_2 \leq \ldots \leq \lambda_n$. For a given $s > 0$ and $C > 0$, let $N(C) = \#\set{k: \lambda_k^s \leq C^2}$. We wish to prove the following result.

\begin{theorem}
	\label{thm:spectral_decay}
	Let $r \to 0$ as $n \to \infty$ sufficiently slowly so that $r(\log n/n)^{1/d} \to \infty$. Then, for each $s > 0$ and $C \leq \sqrt{2}$,
	\begin{equation}
	\label{eqn:spectral_decay}
	N(C) \leq n C^{d/s}.
	\end{equation} 
	with probability tending to one as $n \to \infty$. 
\end{theorem}

\section{Theory}
To prove Theorem \ref{thm:spectral_decay}, we will make use of another graph on points within $\mathcal{D}$, whose spectral properties are very well-understood. Let $\xi$ be the set of evenly spaced lattice points over $\mathcal{D}$; formally $\xi = \set{k/n: k \in [\ell]^d}$ where $\ell = n^{1/d}$, and we define $[\ell]^d = \set{(\ell_1,\ldots,\ell_d): \ell_k \in [\ell] ~\textrm{for each}~ \ell_k}$. Then, let the grid graph over $\xi$ be given by $\wt{G} = (\wt{V},\wt{E})$, where $\wt{V} = \xi$ and $(\xi_k, \xi_{k'})$ is in $\wt{E}$ if $\norm{\xi_k - \xi_{k'}}_1 = 1/\ell$.

Let $\wt{L}$ be the Laplacian of $\wt{G}$, and let $\wt{\lambda_1} \leq \ldots \leq \wt{\lambda}_n$ be the ordered eigenvalues of $\wt{L}$. Write $\wt{N}(C) = \#\set{k: \wt{\lambda}_k^s \leq C^2}$. We have that the desired scaling rate \eqref{eqn:spectral_decay} holds with respect to the grid graph $\wt{G}$.

\begin{lemma}
	\label{lem:spectral_decay_grid}
	For each $s$ and every $C \leq \sqrt{2}$, we have
	\begin{equation}
	\label{eqn:spectral_decay_grid}
	\wt{N}(C) \leq 2^d\big(n C^{d/s} + 1)
	\end{equation}
\end{lemma}
\begin{proof}
	It will be sufficient to show that 
	\begin{equation}
	\label{eqn:spectral_decay_grid_1}
	\lambda_k^s \leq C^2 \Rightarrow \floor{k^{1/d}}^d \leq C^{d/s}n
	\end{equation}
	To show this, observe that for any $\tau \in \mathbb{N}$ and $k = \tau^d$, we have
	\begin{equation*}
	\lambda_k \geq 4 \sin^2 \left(\frac{\pi k^{1/d}}{2n^{1/d}}\right) \geq \frac{\pi^2 k^{2/d}}{4 n^{2/d}} \wedge 2
	\end{equation*}
	Therefore, if $\lambda_k^s \leq C^2$ and $C \leq \sqrt{2}$, this implies
	\begin{equation*}
	\frac{\pi^{2s} k^{2s/d}}{4^s n^{2s/d}} \leq C^{2}
	\end{equation*}
	and rearranging, we obtain
	\begin{equation*}
	k \leq \frac{C^{d/s} 2^d n}{\pi^d} \leq C^{d/s}n
	\end{equation*}	
	If $k^{1/d}$ is not a natural number, applying the same argument to $k' = \floor{k^{1/d}}^d$ yields \eqref{eqn:spectral_decay_grid_1}.
\end{proof}

In light of Lemma \ref{lem:spectral_decay_grid}, to prove Theorem \ref{thm:spectral_decay} it is sufficient to show that $\wt{L} \preceq L$, since by the Courant-Fischer min-max theorem, the ordering $\wt{L} \preceq L$ implies that $\wt{\lambda}_k \leq \lambda_k$ for all $k \in [n]$. The next result details the conditions under which this ordering holds. This condition will be stated with respect to the min-max matching distance between $\xi$ and $X$, i.e. the minimum over all bijections $T: \xi \to X$ such that
\begin{equation*}
\min_{T} \max_{i \in [n]} \abs{T^{-1}(x_i) - x_i} 
\end{equation*}

\begin{lemma}
	\label{lem:partial_ordering_grid}
	For any radius $r$ satisfying
	\begin{equation}
	\label{eqn:radius_condition}
	r \geq 2 \min_{T} \max_{i \in [n]} \abs{T^{-1}(x_i) - x_i} + n^{-1/d}
	\end{equation}
	we have that $\wt{L} \preceq L$. 
\end{lemma}
\begin{proof}
	Let $T_{\star}$ achieve the min-max matching distance, i.e
	\begin{equation*}
	\max_{i \in [n]} \abs{T_{\star}^{-1}(x_i) - x_i} = \min_{T} \max_{i \in [n]} \abs{T^{-1}(x_i) - x_i}.
	\end{equation*}

	It will be sufficient to prove that for every pair $(T_{\star}^{-1}(x_i), T_{\star}^{-1}(x_j)) \in \wt{E}$, the corresponding edge $(i,j)$ is in $E$. To see this, let $A$ denote the adjacency matrix associated with the neighborhood graph $G$, and $\wt{A}$ the adjacency matrix associated with the grid $\wt{G}$. Precisely $A$ is the $n \times n$ matrix with entries $A_{ij} = \eta_r(x_i,x_j)$, and $\wt{A}$ is the $n \times n$ matrix with entries $\wt{A}_{ij} = \1\{T_{\star}^{-1}(x_i), T_{\star}^{-1}(x_j) \in \wt{E} \}$. Our goal is to show that, for every $z \in \Rd$, we have
	\begin{equation*}
	z^T L z = \frac{1}{2}\sum_{i, j = 1}^{n} (z_i - z_j)^2 A_{ij} \leq \frac{1}{2}\sum_{i, j = 1}^{n} (z_i - z_j)^2 \wt{A}_{ij} = z^T \wt{L} z.
	\end{equation*}
	which certainly holds if $\wt{A}_{ij} \leq A_{ij}$ for all $i,j \in [n]$.

	Now, assume $(T_{\star}^{-1}(x_i), T_{\star}^{-1}(x_j)) \in \wt{E}$. This implies
	\begin{align*}
	\norm{x_i - x_j}_2 & \leq \norm{x_i - T_{\star}^{-1}(x_i)}_2 + \norm{T_{\star}^{-1}(x_j) - T_{\star}^{-1}(x_j)}_2 + \norm{x_j - T_{\star}^{-1}(x_j)}_2 \\
	& \leq 2 \max_{i \in [n]} \abs{T_{\star}^{-1}(x_i) - x_i} + n^{-1/d} \\
	& \leq r.
	\end{align*}
	so $\eta_r(x_i,x_j) = 1$ and therefore $(i,j) \in E$. 
\end{proof}

Finally, the following Lemma demonstrates that, for sufficiently large $r$, the condition \eqref{eqn:radius_condition} will hold with high probability.

\begin{lemma}
	\label{lem:matching_distance}
	Assume $P$ has density $p$ which is bounded above and below uniformly over $\D$; that is, there exist constants $p_{\min}$ and $p_{\max}$ such that
	\begin{equation*}
	0 < p_{\min} < p(x) < p_{\max} < \infty, \quad \text{for all $x \in \mathcal{D}$}.
	\end{equation*}
	Then, for any $r = r_n$ such that $r(\log(n)/n)^{1/d} \to \infty$, we have that the event
	\begin{equation*}
	r < 2 \min_{T} \max_{i \in [n]} \abs{T^{-1}(x_i) - x_i} + n^{-1/d}
	\end{equation*}
	occurs with probability tending to $0$ as $n \to \infty$. 
\end{lemma}

Together Lemmas \ref{lem:spectral_decay_grid}, \ref{lem:partial_ordering_grid}, and \ref{lem:matching_distance} imply Theorem \ref{thm:spectral_decay}.
